---
title: "Data mining Course"
author: "Hanjo Odendaal"
date: "27 February 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rmsfuns)
load_pkg(c('dplyr', 'arules', "arulesViz"))
```

# Association Rules

General Format is A -> C    

Can generalize A (the antecedent) to a specific variable or value combination so can apply to various datasets

__Search Heuristic__

Basis of an association analysis algorithm is the generation of frequent itemsets. Is an "apriori algorithm", a generate-and-test type of search algorithm. Only after exploring all of the possibilities of associations containing k items does it consider those contining K + 1 items. For each k, all candidates are tested to determine whether they have enough support.

A frequent itemset is a set of items that occur together frequently enough to be considered as a candidate for generating association rules.

## 3 Measures: Support, Confidence, Lift

__Support__ is a measure of how frequently the  items must appear in the whole dataset before  they can be considered as a candidate asso-  ciation rule.

__Support__ for a collection of items is the proportion of all transactions in which the items appear together

* support(A -> C) = P(A U C)

We use small values of __support__ as are not looking for the obvious ones.  The actual association rules that we retain are those that meet a criterion __confidence__. 

__Confidence__ calculates the proportion of transactions containing A that also contain C.

* confidence(A -> C) = P(C|A) = P(A U C)/P(A)

* confidence(A -> C) = support(A -> C)/support(A)

Typically looking for larger values of confidence

__Lift__ is the increased likelihood of C being in a transaction if A is included in the transaction:

* lift(A -> C) = confidence(A -> C)/support(C)

Another measure: __Leverage__, which captures the fact that a higher frequency of A and C with a lower lift may be interesting:

* leverage(A->C) = support(A->C)-support(A) \* support(C)

## Titanic dataset

```{r}
data("Titanic")
df <- as.data.frame(Titanic)

# make new object to hold data
titanic.raw <- NULL

# restructure it as a character matrix
for(i in 1:4) {
   titanic.raw <- cbind(titanic.raw, rep(as.character(df[,i]), df$Freq))
}

titanic.raw <- as.data.frame(titanic.raw)
names(titanic.raw) <- names(df)[1:4]

rules.all <- apriori(titanic.raw)

rules.all

inspect(rules.all)
```

As a common phenomenon for association rule mining, many rules generated above are uninteresting. Suppose that we are interested in only rules with rhs indicating survival, so we set rhs=c("Survived=No", "Survived=Yes") in appearance to make sure that only "Survived=No" and "Survived=Yes" will appear in the rhs of rules. All other items can appear in the lhs, as set with default="lhs". 

In the above result rules.all, we can also see that the left-hand side (lhs) of the first rule is empty. To exclude such rules, we set minlen to 2 in the code below. Moreover, the details of the progress are suppressed with verbose=F. 

After association rule mining, rules are sorted by lift to make high-lift rules appear first.

```{r}
rules <- apriori(titanic.raw, control = list(verbose=F),
                 parameter = list(minlen=2, supp=0.005, conf=0.8),
                 appearance = list(rhs=c("Survived=No", 
                                         "Survived=Yes"),
                                   default="lhs"))

quality(rules) <- round(quality(rules), digits=3)
rules.sorted <- sort(rules, by="lift")

inspect(rules.sorted)
```

When other settings are unchanged, with a lower minimum support, more rules will be produced, and the associations between itemsets shown in the rules will be more likely to be by chance. In the above code, the minimum support is set to 0.005, so each rule is supported at least by 12 (=ceiling(0.005 * 2201)) cases, which is acceptable for a population of 2201. 

__Support__, __confidence__ and __lift__ are three common measures for selecting interesting association rules. Besides them, there are many other interestingness measures, such as 
* chi-square
* conviction
* gini
* leverage

More than 20 measures can be calculated with function `interestMeasure()` in the arules package.

```{r}
interestMeasure(rules.sorted,  transactions = titanic.raw)

quality(rules) <- 
  interestMeasure(rules.sorted,  transactions = titanic.raw)

data("Income")
rules <- apriori(Income)

## calculate a single measure and add it to the quality slot
quality(rules) <- cbind(quality(rules), 
                        hyperConfidence = interestMeasure(rules, measure = "hyperConfidence", 
                                                          transactions = Income))

inspect(head(rules, by = "hyperConfidence"))
```

Some rules generated in the previous section (see rules.sorted, page 89) provide little or no extra information when some other rules are in the result. 

For example, the above rule 2 provides no extra knowledge in addition to rule 1, since rules 1 tells us that all 2nd-class children survived.  Generally speaking, when a rule (such as rule 2) is a super rule of another rule (such as rule 1) and the former has the same or a lower lift, the former rule (rule 2) is considered to be redundant.  Other redundant rules in the above result are rules 4, 7 and 8, compared respectively with rules 3, 6 and 5. 

Below we prune redundant rules. Note that the rules have already been sorted descendingly by lift.

To sum up redundancy, if we see that the confidence of Class = 1st, Sex = Female is 1, then by adding if its an adult or an infant does not matter as it does not add information

```{r}
### Just make 
prune_arule <- function(rules, verbose = T) {
  
  rules <- sort(rules, by="lift")
  subset.matrix <- is.subset(rules.sorted, rules.sorted)
  subset.matrix[lower.tri(subset.matrix, diag = T)] <- NA
  redundant <- colSums(subset.matrix, na.rm = T) >= 1
  if(verbose)
  cat("Removing rules:\n", 
      which(redundant), "\n",
      paste(names(which(redundant)),"\n"))
  
  rules[!redundant]
}

rules.pruned <- prune_arule(rules)

inspect(rules.pruned)
```

__While it is easy to find high-lift rules from data, it is not an easy job to understand the identified rules__. It is not uncommon that the association rules are misinterpreted to find their business meanings.  

For instance, in the above rule list `rules.pruned`, the first rule `{Class=2nd, Age=Child} => {Survived=Yes}` has a confidence of one and a lift of three and there are no rules on children of the 1st or 3rd classes. Therefore, it might be interpreted by users as children of the 2nd class had a higher survival rate than other children. __This is wrong!__ 

* The rule states only that all children of class 2 survived, but provides no information at all to compare the survival rates of different classes. 

To investigate the above issue, we run the code below to find rules whose rhs is `Survived=Yes` and lhs contains `Class=1st`, `Class=2nd`, `Class=3rd`, `Age=Child` and `Age=Adult` only, and which contains no other items (default=`none`). We use lower thresholds for both support and confidence than before to find all rules for children of different classes.

* Remember to also change `supp` and `conf` as we work with different samples in the data
* This will give us a much clearer view of the proportional survival rates between classes

```{r}
# create a function to run a truncated apriori - work out supp and conf - input list of rhs and lhs variables
# function(trunc_list, supp, conf)

rules <- apriori(titanic.raw, 
                 parameter = list(minlen=3, supp=0.002, conf=0.2),
                 appearance = list(rhs=c("Survived=Yes"),
                                   lhs=c("Class=1st", "Class=2nd", "Class=3rd",
                                         "Age=Child", "Age=Adult"),
                                   default="none"), 
                 control = list(verbose=F))

rules.sorted <- rules %>% sort(by="confidence")

inspect(rules.sorted)
```

# Visualizing arules

Here we show some ways to visualize association rules, including scatter plot, balloon plot, graph and parallel coordinates plot. More examples on visualizing association rules can be found in the vignettes of package arulesViz [Hahsler and Chelluboina, 2012] which is included in your folder of class materials for today.

```{r}
library(arulesViz)

# figure 9.1 in PDF
plot(rules.all) 

# figure 9.2 in PDF
plot(rules.all, method="grouped")

# figure 9.3 in PDF
plot(rules.all, method="graph")

# figure 9.4
plot(rules.all, method="graph", control=list(type="items"))

# figure 9.5
plot(rules.all, method="paracoord", control=list(reorder=TRUE))


plot(rules.all, 
     method = "matrix", 
     measure = c("lift", "confidence"), 
     control = list(reorder=TRUE))

plot(rules, 
     method = "matrix3D", 
     measure = "lift", 
     control = list(reorder=TRUE))
```

## Groceries

```{r}
# Groceries contains sales data of groceries
# with 9835 transactions and 169 items.
data("Groceries")
summary(Groceries)

# Whole milk most popular. Average transaction
# has less than 5 items.

# We mine association rules using
# Apriori algorithm implemented in arules

rules <- apriori(Groceries,
                 parameter=list(support=0.001,
                                confidence=0.5))

rules %>% 
  sort(by = "lift") %>% 
  head(3) %>% 
  inspect
```

Here we see that when people buy `Instant food products, soda`, it is ~19 more likely (`lift`) on average that they will buy `hamburger meat`

Generic Plot Syntax is:
```{r}
plot(x, method = NULL, measure = "support",
            shading = "lift", interactive = FALSE,
            data = data_set, control = list(reorder=TRUE))
```

* x = set of rules to be visualized
* method = visualization method
* measure and shading = interest measures for plot
* interactive = if we want to interactively explore or just present the rules
* data = transaction data set used to mine rules control = list with further control arguments.

```{r}
plot(rules)

rules %>% quality %>% head

rules %>% 
  plot(., 
     measure=c("support", "lift"), 
     shading="confidence")
```

In a "two-key plot", support and confidence are used for the x and y-axes and the color of the points is used to indicate "order" which is the number of items contained in the rule.

```{r}
rules %>% 
  plot(., 
       shading="order", 
       control=list(main="Two-key plot"))
```

